---
title: "Practical Machine Learning Project"
author: "Joseph Coggins"
date: "September 26, 2015"
output: html_document
---
#Introduction

This is homework assignment for Coursera course Practical Machine Learning.
The course is provided by John Hopkins University.  

The scripts used were write on a PC, using RStudio Version 0.99.484.

My GitHub repos are located at: https://github.com/jcoggins

This assignment is located at: https://github.com/jcoggins/Practical-Machine-Learning-Project/blob/master/ML_Project_JCoggins.html



##Background:

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

The machine learning package is the caret package. The ggplot2
package is a good standardized plotting package.
```{r}
library(ggplot2)
library(caret)
set.seed(34323)

```
##Data Sources

Machine training source: "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

Test data on machine source: "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

##Goal
The purpose of the project is to train a machine with an accuracy of prediction
of greater than 90%.  The classe variable is used for the prediction.  Cross-validation
and out of sample error should be shown. The predication model choose should successfully
predict the classe values for the 20 possible submissions.

Here is the URL path to the training.csv file
```{r}
trainingURL = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
Here is the URL path to the testing.csv file
```{r}
testingURL = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Extracting the csv files
Now we can read the csv files into memory
training
```{r}
training = read.csv(url(trainingURL),, header=T, sep=",", na.strings=c("NA", "#DIV/0!", ""))
```
testing
```{r}
testing = read.csv(url(testingURL), header=T, sep=",", na.strings=c("NA", "#DIV/0!", "")) 
```
Exploratory Analysis:
```{r}
#summary(training, 5)
#head(training, 5)
#names(training)
```
The head() without using na.strings shows NA, blanks, #Div/0! and near zero data.
This data should be cleaned from the training data
to train the machine on clean data. Colums without
data should be removed as well.

Let's Partition the given Training Data
```{r}
inTrain = createDataPartition(y=training$classe, p=0.7, list=FALSE)
trainingData = training[inTrain,]
testingData = training[-inTrain,]
dim(trainingData)
dim(testingData)
```

Determine the datatype for each column
```{r}
#sapply(trainingData, class)
```
Using Google search the correct link was found at
Human Activity Recognition - Groupware@LES - PUC-Rio
groupware.les.inf.puc-rio.br/har
Human Activity Recognition - HAR - has emerged as a key research area in the ... dataset 

##Cleaning Data
Cleaning out columns with near zero variance
```{r}
nzv = nearZeroVar(trainingData, saveMetrics = TRUE)
trainingData = trainingData[,nzv$nzv==FALSE]

nzv = nearZeroVar(testingData, saveMetrics = TRUE)
testingData = testingData[,nzv$nzv==FALSE]

#head(trainingData)
```
##Data Transformation
Create another subset
```{r}
trainingData = trainingData[c(-1)]

trainingDataA = trainingData
for(i in 1:length(trainingData)) {
  if(sum(is.na(trainingData[,i])) /nrow(trainingData) >= 0.60) {
    for(j in 1:length(trainingDataA)) {
      if(length( grep(names(trainingData[i]), names(trainingDataA)[j])) == 1){
        trainingDataA = trainingDataA[, -j]
      }
    }
  }
}

trainingData = trainingDataA
dim(trainingData)
```
##Cleaning and Transforming Test Data
Cleanup the testingData and testing data sets
```{r}
removeA = colnames(trainingData)  # 60% column removal
removeB = colnames(trainingData[, -58]) # removes the classe column (last column)
testingData = testingData[removeA]
testing = testing[removeB]
dim(testingData)
dim(testing)
```
##Coercing the Data
Coerce the data into the same type
```{r}
for(i in 1:length(testing)) {
  for(j in 1:length(trainingData)) {
    if(length(grep(names(trainingData[i]), names(testing)[j]))== 1) {
      class(testing[j]) = class(trainingData[i])
    }
  }
}

testing = rbind(trainingData[2, -58], testing)
testing = testing[-1,]
```
## First Plot
Data Plot
```{r, echo=FALSE}
library(ggplot2)
table(trainingData$classe)
qplot(roll_forearm, roll_arm, colour=classe, data=trainingData)

```

## Try Decision Tree
Decision Tree Plot
```{r, echo=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
modFit = rpart(classe ~., data=trainingData, method="class")
library(rattle)
fancyRpartPlot(modFit)

```

Prediction:
```{r}
prediction = predict(modFit, testingData, type="class")
table(prediction, testingData$classe)
```

Use confusion matrix to check Accuracy
```{r}
cmTree = confusionMatrix(prediction, testingData$classe)
cmTree
```

Plot Decision Tree CM
```{r, echo=FALSE}
plot(cmTree$table, col=cmTree$byClass, main=paste("Decision Tree CM: Accuracy = ", round( cmTree$overall['Accuracy'], 4)))
```

## Try Random Forests
Random Forests Prediction
```{r}
library(randomForest)
  modFitRF = randomForest(classe ~ ., data=trainingData)
  predictionRF = predict(modFitRF, testingData, type="class")
  cmRF = confusionMatrix(predictionRF, testingData$classe)
  cmRF
```
Plot Model Fit
```{r, echo=FALSE}
plot(modFit)
```
Plot Random Forests
```{r, echo=FALSE}
plot(cmRF$table, col=cmTree$byClass, main=paste("Random Forest CM: Accuracy = ", round( cmRF$overall['Accuracy'], 4)))
```
## Generating Submission Files based on the Prediction from Random Forests

Generating Files for Submission
```{r}
  predictionRF = predict(modFitRF, testing, type="class")
  
  pml_write_files = function(x) {
    n = length(x)
    for( i in 1:n){
      filename = paste0("problem_id_", i, ".txt")
      write.table(x[i], file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
    }
  }
  
  pml_write_files(predictionRF)

```
